Combining living and fossil taxa into phylogenies: the missing data issue

### 2014-05-28 Manuscript/TEM_draft.tex
Added statistical comparisons and pre results

%Abstract
\subsubsection{Abstract}
Living species represent less than 1\% of all species that have ever lived. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology.
This fact has led to an increasing consensus among scientists that both fossil and living taxa must be included in macroevolutionary studies.
One approach, the total evidence approach, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both fossil and living taxa at the tips.
Although the total evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

@@

@@

\section{Methods} %no material needed?
To explore the effect of missing data on total evidence trees topologies we used the following protocol (note that we explain each step in detail below this general outline -Fig. ~\ref{Fig_Outline}):
\begin{enumerate}
\item{Generating the matrix:} \\
\label{step:generate_matrix}
We randomly generated birth death tree (hereafter called the "true" tree, Table ~\ref{Tab_glossary}) to infer a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the "complete" matrix, Table ~\ref{Tab_glossary}).
\item{Removing data:} \\
\label{step:remove_data}
We removed data from the morphological part of the "complete" matrix to simulate the effects of missing data by modifying three parameters (1) the proportion of missing living taxa ($M_{L}$), (2) the proportion of missing data in the fossil taxa ($M_{F}$) and (3) the proportion of missing morphological characters ($M_{C}$) (the resulting matrices are called hereafter the "missing-data" matrix, Table ~\ref{Tab_glossary}).
\item{Building phylogenies:} \\
\label{step:build_phylo}
We inferred Bayesian phylogenetic trees from the "complete" matrix and from the "missing-data" matrices resulting in respectively a tree generate from a matrix containing no missing data in the morphological part (hereafter called the "best" tree, Table ~\ref{Tab_glossary}) and trees inferred from matrices with missing data in their morphological part (hereafter called the "missing-data" trees, Table ~\ref{Tab_glossary}).
\item{Comparing topologies:} \\
\label{step:compare_topo}
We then compared the "best" tree to the "missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on the topologies of the phylogenies we estimated.
\end{enumerate}
To measure the effect of missing data distribution, we repeated steps ~\ref{step:generate_matrix} to ~\ref{step:compare_topo} with the exact same fixed parameters 51 (3$\times$17) times. %that's because I had 3 cluster accounts available...
A list of all the terms used in this paper is available in table ~\ref{Tab_glossary}.

@@

@@

@@

@@

%
%
%

@@

\subsubsection{Triple distance ($T_{x,y}$) \citep{dobson1975triplets}}
This metric measures the number of different subtrees of three clades between two given trees.
Each triplet can be written as $I_{ijk}$=(\textit{ijk}).
Where $I_{ijk}$ is equal to 0 if the the two triplets (\textit{ijk}) are the same in the two trees otherwise $I_{ijk}$ is equal to 1.
For any rooted binary tree there are only three possible combinations per triplets: ((\textit{j},\textit{k}),\textit{i});, ((\textit{i},\textit{k}),\textit{j}); and ((\textit{i},\textit{j}),\textit{k}); \citep{johnson1998}.
If the trees used are not fully binary, a fourth triplet combination is possible: (\textit{i},\textit{j},\textit{k});.
One can calculate $S_n$, the triplet distance between two trees as:
\begin{equation}
S_n = \sum_{ijk} I_{ijk}
\end{equation}
Where:
\begin{equation}
\sum_{ijk} = \binom{n}{4} = \frac{n!}{4!(n-4)!}
\end{equation}
And where \textit{n} is the number of taxa in both trees (modified from \citet{critchlowthe1996}).
If $S_n$=0, the trees are the same (i.e. no taxa as been displaced). When $S_n$ = $\binom{n}{4}$, the trees are the most different possible (i.e. every taxa as been displaced).
This metric therefore illustrates the amount of displaxed taxa but is less sensitive to the placement of individual taxa and to taxa of highly uncertain placement (e.g. fossil taxa) than the Robinson-Foulds distance \citep{critchlowthe1996,johnson1998,wiensmissing2003} and can therefore be used as a proxy to estimate the robustness of the tree to flying taxa (see \hyperref[supplementaries]{supplementaries}).

@@

@@

@@

\section{Results}
%Do the check for L00F00C00
\subsection{Building the trees}
Bayesian chains converged (SD $<$ 0.01 and ESS $>>$ 200 for each tree).
\subsection{Comparing topologies}
Did it worked? See sub-matrices
\subsection{Effect of $M_L$}
In ML and in Bayesian
\subsection{Effect of $M_F$}
In ML and in Bayesian
\subsection{Effect of $M_C$}
In ML and in Bayesian
\subsection{Combined effect}
In ML and in Bayesian

@@

@@

@@

\section{Acknowledgements}
We would like to thank Trevor Hodkinson %Also Gavin Thomas, Emmanuel Douzery and Frédéric Delsuc if not authors.
for his useful comments on the simulation protocol and Paddy Doyle for the assistance on using the computer cluster.
All calculations were performed on the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing.
This cluster was funded through grants from Science Foundation Ireland.

@@

\subsubsection{Random Pairwise Bayesian Tree Comparison (RPBTC)}
We assessed the power of the Random Pairwise Bayesian Tree Comparison (RPBTC) method by comparing 1000 random trees from a posterior distribution trees set to another 1000 random trees from the same posterior distribution trees set.
We repeated this 100 times independently using the same posterior distribution trees set each time resulting in 100 replicates of the same posterior distribution trees set compared 1000 times.
We used an anova to test if there was no significant difference between the replicates so that the RBTC can be replicated.
We applied this protocol on a poorly resolved tree (Low Score), a resolved tree with low support value (Medium Score) and a resolved tree with high support values (High Score).
Results are available in table ~\ref{RPBTC_testing}.

\begin{table}[ht]
\caption{Anova results: difference between 100 replicates using the RPBTC method}
\centering
\begin{tabular}{rllrrrr}
  \hline
  Tree.Type & Used.metric & Replicates & Df & F.value & p.value \\ 
  \hline
  Low Score & RF & 100.00 & 99.00 & 0.74 & 0.98 \\ 
  Low Score & Tr & 100.00 & 99.00 & 0.97 & 0.58 \\ 
  Medium Score & RF & 100.00 & 99.00 & 0.64 & 1.00 \\ 
  Medium Score & Tr & 100.00 & 99.00 & 0.45 & 1.00 \\ 
  High Score & RF & 100.00 & 99.00 & 0.20 & 1.00 \\ 
  High Score & Tr & 100.00 & 99.00 & 0.37 & 1.00 \\ 
  \hline
\end{tabular}
\end{table}
\label{RPBTC_testing}

>>>>

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\subsubsection{Abstract}
Living species represent less than 1\% of all species that have ever lived. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology.
This fact has led to an increasing consensus among scientists that both fossil and living taxa must be included in macroevolutionary studies.
One approach, the total evidence approach, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both fossil and living taxa at the tips.
Although the total evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------

@@

%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------

\section{Methods} %no material needed?
To explore the effect of missing data on total evidence trees topologies we used the following protocol (note that we explain each step in detail below this general outline -Fig. ~\ref{Fig_Outline}):
\begin{enumerate}
\item{Generating the matrix:} \\
\label{step:generate_matrix}
We randomly generated birth death tree (hereafter called the "true" tree, Table ~\ref{Tab_glossary}) to infer a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the "complete" matrix, Table ~\ref{Tab_glossary}).
\item{Removing data:} \\
\label{step:remove_data}
We removed data from the morphological part of the "complete" matrix to simulate the effects of missing data by modifying three parameters (1) the proportion of missing living taxa ($M_{L}$), (2) the proportion of missing data in the fossil taxa ($M_{F}$) and (3) the proportion of missing morphological characters ($M_{C}$) (the resulting matrices are called hereafter the "missing-data" matrix, Table ~\ref{Tab_glossary}).
\item{Building phylogenies:} \\
\label{step:build_phylo}
We inferred Bayesian phylogenetic trees from the "complete" matrix and from the "missing-data" matrices resulting in respectively a tree generate from a matrix containing no missing data in the morphological part (hereafter called the "best" tree, Table ~\ref{Tab_glossary}) and trees inferred from matrices with missing data in their morphological part (hereafter called the "missing-data" trees, Table ~\ref{Tab_glossary}).
\item{Comparing topologies:} \\
\label{step:compare_topo}
We then compared the "best" tree to the "missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on the topologies of the phylogenies we estimated.
\end{enumerate}
To measure the effect of missing data distribution, we repeated steps ~\ref{step:generate_matrix} to ~\ref{step:compare_topo} with the exact same fixed parameters 51 (3$\times$17) times. %that's because I had 3 cluster accounts available... If it's a too odd number, no worries, I'll just remove the last chain.
A list of all the terms used in this paper is available in table ~\ref{Tab_glossary}.

@@

%---------------------------------------------
%       Generating the matrix
%---------------------------------------------

@@

%---------------------------------------------
%       Removing data
%---------------------------------------------

@@

%---------------------------------------------
%       Building phylogenies
%---------------------------------------------

@@

%---------------------------------------------
%       Comparing Topologies
%---------------------------------------------

@@

\subsubsection{Triple distance ($T_{x,y}$) \citep{dobson1975triplets}}
This metric measures the number of different subtrees of three clades between two given trees.
Each triplet can be written as $I_{ijk}$=(\textit{ijk}).
Where $I_{ijk}$ is equal to 0 if the the two triplets (\textit{ijk}) are the same in the two trees otherwise $I_{ijk}$ is equal to 1.
For any rooted binary tree there are only three possible combinations per triplets: ((\textit{j},\textit{k}),\textit{i});, ((\textit{i},\textit{k}),\textit{j}); and ((\textit{i},\textit{j}),\textit{k}); \citep{johnson1998}.
If the trees used are not fully binary, a fourth triplet combination is possible: (\textit{i},\textit{j},\textit{k});.
One can calculate $S_n$, the triplet distance between two trees as:
\begin{equation}
S_n = \sum_{ijk} I_{ijk}
\end{equation}
Where:
\begin{equation}
\sum_{ijk} = \binom{n}{4} = \frac{n!}{4!(n-4)!}
\end{equation}
And where \textit{n} is the number of taxa in both trees (modified from \citet{critchlowthe1996}).
If $S_n$=0, the trees are the same (i.e. no taxa as been displaced). When $S_n$ = $\binom{n}{4}$, the trees are the most different possible (i.e. every taxa as been displaced).
This metric therefore illustrates the amount of displayed taxa but is less sensitive to the placement of individual taxa and to taxa of highly uncertain placement (e.g. fossil taxa) than the Robinson-Foulds distance \citep{critchlowthe1996,johnson1998,wiensmissing2003} and can therefore be used as a proxy to estimate the robustness of the tree to flying taxa (see \hyperref[supplementaries]{supplementaries}).

@@

\subsubsection{Statistical difference between tree topologies}
To assess the difference between the tree topologies, whether they were inferred in a Bayesian or a ML framework, we performed a group comparison test for each set of configurations (e.g. effect of $M_L$ in Bayesian inference, combined effect of $M_F$ and $M_C$ in ML inference, etc...).
We used the pairwise tree comparisons as factors (i.e. the "Best" tree vs. one of each of the 125 "missing-data" trees) and the values of the 51 replicates from the pairwise comparison as a response variable.
The 51 replicates values were either the replicates of the 51 pairwise tree comparisons in ML or the modes of the 51 RPBTC posterior distributions in Bayesian.
When we found a significant difference between the groups, we performed a pairwise comparison test to analyse which pairs of groups differ from the others.
Depending on whether the data was normal with variance homoscedasticity or not (controlled by performing respectively a shapiro.test and a a bartlett.test in R\{stats\}), we used respectively parametric or non-parametric tests for the group comparison and the pairwise comparisons as suggested by \citet{ruxtontime2008} (Table ~\ref{Stats_test}). %link broken

\begin{table}[ht]
\caption{Statistical tests}
\centering
\begin{tabular}{lrll}
  \hline
   & test & statistics & code\{package\} \\ [0.5ex]
  \hline
  Groups comparisons & ANOVA & parametric & anova\{stats\} \\ [0.5ex]
                     & Kruskal-Wallis & non-parametric & kruskal.test\{stats\} \\[0.5ex]
  \hline
  Pairwise comparisons & Tukey HSD & parametric & TukeyHSD\{stats\} \\ [0.5ex]
                       & Nemenyi-Damico-Wolfe-Dunn & non-parametric & kruskalmc\{pgirmess\} \\ [0.5ex]
  \hline
\end{tabular}
\end{table}
\label{Stats_test}

@@

%---------------------------------------------
%       Empirical data - TO DO
%---------------------------------------------

@@

%---------------------------------------------
%
%       RESULTS
%
%---------------------------------------------

\section{Results}
%Do the check for L00F00C00
\subsection{Building the trees}
Bayesian chains converged (SD $<$ 0.01 and ESS $>>$ 200 for each tree).
Good a quick convergence when using the sub-matrices especially for the "horizontal" sub-matrices.
Good enough phylogenetic signal.
\subsection{Comparing topologies}
Did it worked? See sub-matrices

\subsection{Effect of the missing data parameters}

\subsubsection{Effect of $M_L$}
In ML...

Topological recovery was bad in Bayesian, the NTS score was only slightly above 0 which means that the mode of the RPBTC was only slightly different than expected by chance.
However, it seems that there is a small difference depending on the metric used.
For the Triple metric there is no significant effect of the $M_L$ parameter, in other words, if living taxa are removed, there is no effect on the placement of "flying" taxa.
For the Robinson-Fould metric, there is a significant effect of the $M_L$ leading small decrease in topological recovery when $M_L$ increases.
In other words, if living taxa are removed, clades are less conserved.
However, in both cases, the effect of the parameter $M_L$ is still minimal compared to trees generated in a ML framework since the NTS score was only slightly higher than expected by chance.

%Tables model...

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Wed May 28 17:54:17 2014
\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
  \hline
 & Kruskal.Wallis.chi.squared & df & p.value \\ 
  \hline
1 & 65.62 & 4.00 & 0.00 \\ 
   \hline
\end{tabular}
\end{table}

% latex table generated in R 3.0.2 by xtable 1.7-1 package
% Wed May 28 17:54:17 2014
\begin{table}[ht]
\centering
\begin{tabular}{rrrl}
  \hline
 & obs.dif & critical.dif & difference \\ 
  \hline
1-2 & 45.87 & 41.00 & TRUE \\ 
  1-3 & 60.33 & 41.00 & TRUE \\ 
  1-4 & 76.83 & 41.00 & TRUE \\ 
  1-5 & 99.80 & 41.00 & TRUE \\ 
  2-3 & 14.46 & 41.00 & FALSE \\ 
  2-4 & 30.96 & 41.00 & FALSE \\ 
  2-5 & 53.93 & 41.00 & TRUE \\ 
  3-4 & 16.50 & 41.00 & FALSE \\ 
  3-5 & 39.47 & 41.00 & FALSE \\ 
  4-5 & 22.97 & 41.00 & FALSE \\ 
   \hline
\end{tabular}
\end{table}

\subsubsection{Effect of $M_F$}
In ML...

Similarly than for the effect of $M_L$, Bayesian tree recovery was only slightly better than expected by chance.
And in the same way as described above, there seems to be a small difference depending on the metric used.
There is no significant effect of the $M_F$ parameter when using the Triplet metric but the effect is significant when using the Robinson-Fould metric.
However, as mentioned previously the effect of the parameter $M_F$ is still minimal in a Bayesian framework.

\subsubsection{Effect of $M_C$}
In ML...

The effect of $M_C$ in Bayesian framework is similar as the one described for $M_L$ and $M_F$: tree recovery is only slightly better than expected by chance and only the Robinson-Fould metric shows a significant effect of the $M_C$ parameter on tree recovery.
\subsubsection{Combined effect}

In ML...

In Bayesian, same pattern as above.
No particular effect of data combination, but more an effect of the overall amount of data.

@@

%---------------------------------------------
%
%       DISCUSSION
%
%---------------------------------------------

@@

%Potential reviewers comments?
%What about it when you vary the proportion of fossil taxa?

%---------------------------------------------
%
%       CONCLUSION
%
%---------------------------------------------

@@

%---------------------------------------------
%
%       ...
%
%---------------------------------------------

\section{Acknowledgements}
We would like to thank Trevor Hodkinson and Andrew Jackson %Also Gavin Thomas, Emmanuel Douzery and Frédéric Delsuc if not authors.
for his useful comments on the simulation protocol and Paddy Doyle for the assistance on using the computer cluster.
All calculations were performed on the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing.
This cluster was funded through grants from Science Foundation Ireland.

@@

\subsubsection{Random Pairwise Bayesian Tree Comparison (RPBTC)}
We assessed the power of the Random Pairwise Bayesian Tree Comparison (RPBTC) method by comparing 1000 random trees from a posterior distribution trees set to another 1000 random trees from the same posterior distribution trees set.
We repeated this 100 times independently using the same posterior distribution trees set each time resulting in 100 replicates of the same posterior distribution trees set compared 1000 times.
We used an anova to test if there was no significant difference between the replicates so that the RBTC can be replicated.
We applied this protocol on a poorly resolved tree (Low Score), a resolved tree with low support value (Medium Score) and a resolved tree with high support values (High Score).
Results are available in table ~\ref{RPBTC_testing}. %link broken

\begin{table}[ht]
\caption{Group comparison results: difference between 100 replicates using the RPBTC method} %test with TreeCmp.anova
\centering
\begin{tabular}{rllrrrr}
  \hline
  Tree.Type & Used.metric & Replicates & Df & F.value & p.value \\ 
  \hline
  Low Score & RF & 100.00 & 99.00 & 0.74 & 0.98 \\ 
  Low Score & Tr & 100.00 & 99.00 & 0.97 & 0.58 \\ 
  Medium Score & RF & 100.00 & 99.00 & 0.64 & 1.00 \\ 
  Medium Score & Tr & 100.00 & 99.00 & 0.45 & 1.00 \\ 
  High Score & RF & 100.00 & 99.00 & 0.20 & 1.00 \\ 
  High Score & Tr & 100.00 & 99.00 & 0.37 & 1.00 \\ 
  \hline
\end{tabular}
\end{table}
\label{RPBTC_testing}
